---
title: "bulk-org figures (AB) V1.1"
author: "Raphael Kubler"
date: "01/24/2024"
output:
  html_document:
    toc: yes
    df_print: paged
  rmarkdown::html_document:
    theme: united
    highlight: tango
    code_folding: hide
    toc: yes
    toc_float: yes
    df_print: paged
    smooth_scroll: yes
    number_sections: no
    self_contained: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

load('~/Documents/R projects/bulk-org/01152024_AB figures analysis.RData')


library(tibble)
library(matrixStats)
library(factoextra)
library(broom)
library(ggpubr)
library(tidyr)
library(dplyr)
library(plyr)
library(ggeasy)
library(ggrepel)
library(gridExtra)
library(ComplexHeatmap)
library(DESeq2)
library(dplyr)
library(plyr)
library(clusterProfiler)
library(tidyr)
library(biomaRt)
library(variancePartition)
library(ggrastr)
library(kableExtra)
```

# CO vs COiMG in control samples

## PCA plots
```{r pca, echo=TRUE, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}
#pca plots

pca.ctr[pca.ctr$Line %in% 'MSN38',]$Line <- 'MSN-38'
pca.ctr[pca.ctr$Line %in% 'MSN9',]$Line <- 'MSN-9'
pca.ctr[pca.ctr$Line %in% 'WTC11',]$Line <- 'WTC-11'

pca.ctr$Organoids <- NA
pca.ctr[pca.ctr$COiMg %in% 'yes',]$Organoids <- 'COiMG'
pca.ctr[pca.ctr$COiMg %in% 'no',]$Organoids <- 'CO'


PCAplot(pca.ctr, "Organoids", Shape = "Line", PoV.df=PoV.ctr, pc.1 = 1, pc.2 = 2, colors=c('lightgrey','black'))


```

## Volcano plot

```{r volcano.new_coimg, echo=TRUE, fig.height = 4, fig.width = 6, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}

volcano_plot(data.frame(new_coimg), title = 'CO vs COiMg',
             annotate_by=unlist(AB.genes), ymax1 = 30, ymax2 = 31, xmax1 = -5, xmax2 = 8) + 
  scale_x_continuous(breaks = c(-5,-3,-1, 0,1,3,5),
                     labels = c("-5",
                                '-3',
                                "-1",
                                "0",
                                "1",
                                '3',
                                "5"))

```

## Boxplot genes of interest

```{r dea, echo=TRUE, fig.height = 4, fig.width = 7, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}
meta.ctr$Organoids <- NA
meta.ctr[meta.ctr$COiMg %in% 'yes',]$Organoids <- 'COiMg'
meta.ctr[meta.ctr$COiMg %in% 'no',]$Organoids <- 'CO'

boxplot1 <- data.frame('Sample'=rownames(meta.ctr),'Organoids'=meta.ctr$Organoids,t(batch.ctr[rownames(batch.ctr) %in% c('C3',"C1QA","CX3CR1","TLR4"),]))
boxplot2 <- data.frame('Sample'=rownames(meta.ctr),'Organoids'=meta.ctr$Organoids,t(batch.ctr[rownames(batch.ctr) %in% c("CD68","TREM2",'TYROBP',"SYK"),]))
boxplot3 <- data.frame('Sample'=rownames(meta.ctr),'Organoids'=meta.ctr$Organoids,t(batch.ctr[rownames(batch.ctr) %in% c("SPP1","CSF1R","IRF8"),]))

# Reshape the data using gather
expression_1 <- gather(boxplot1, key = "Gene", value = "Expression", -Sample, -Organoids)
expression_2 <- gather(boxplot2, key = "Gene", value = "Expression", -Sample, -Organoids)
expression_3 <- gather(boxplot3, key = "Gene", value = "Expression", -Sample, -Organoids)


ggplot(expression_1,aes(x = Gene, y = Expression, fill = Organoids)) +
  geom_boxplot(position = position_dodge(width = 0.8), width = 0.3, alpha = 0.8, outlier.shape = NA)  +
  geom_point(aes(fill = Organoids), size = 3, shape = 21, position = position_jitterdodge(dodge.width = 0.8, jitter.height = 0.2, jitter.width = 0.01)) +
  theme_minimal() + 
  scale_fill_manual(values=c("white","black")) +
  labs(y = "Standardized expression")


ggplot(expression_2,aes(x = Gene, y = Expression, fill = Organoids)) +
  geom_boxplot(position = position_dodge(width = 0.8), width = 0.3, alpha = 0.8, outlier.shape = NA)  +
  geom_point(aes(fill = Organoids), size = 3, shape = 21, position = position_jitterdodge(dodge.width = 0.8, jitter.height = 0.2, jitter.width = 0.01)) +
  theme_minimal() + 
  scale_fill_manual(values=c("white","black")) +
  labs(y = "Standardized expression")


ggplot(expression_3,aes(x = Gene, y = Expression, fill = Organoids)) +
  geom_boxplot(position = position_dodge(width = 0.8), width = 0.3, alpha = 0.8, outlier.shape = NA)  +
  geom_point(aes(fill = Organoids), size = 3, shape = 21, position = position_jitterdodge(dodge.width = 0.8, jitter.height = 0.2, jitter.width = 0.01)) +
  theme_minimal() + 
  scale_fill_manual(values=c("white","black")) +
  labs(y = "Standardized expression")


```


## Patir microglia genes heatmap

```{r heatmap.top50, echo=TRUE, fig.height = 10, fig.width = 6, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}

ra.ctr <- HeatmapAnnotation(
  Organoids = meta.ctr$Organoids,
  col = list(
    Organoids = c("COiMg"='black','CO'='white')),
  show_annotation_name = T,
  show_legend = T)


ComplexHeatmap::Heatmap(t(hm_counts.ctr),
                        cluster_rows = T,
                        cluster_columns = T,
                        show_row_dend = T,
                        show_column_dend = T,
                        show_row_names = F,
                        show_column_names = F,
                        top_annotation = ra.ctr,
                        right_annotation = rowAnnotation(
                          text = anno_text(colnames(hm_counts.ctr), rot = 0, just = "left",gp=gpar(fontsize=8))),
                        col = colorRampPalette(rev(RColorBrewer::brewer.pal(9,"RdBu")))(10),
                        name = "Z-score expression")


```
## Enrichment heatmap (gene sets of interest)
```{r coimg.table, echo=TRUE, fig.height = 6, fig.width = 5, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}

ComplexHeatmap::Heatmap(DEG.enrich_coimg[!rownames(DEG.enrich_coimg) %in% 'Lipid metabolism',],
                        cluster_rows = F,
                        cluster_columns = F,
                        show_row_dend = F,
                        show_column_dend = F,
                        show_row_names = T,
                        col = colorRampPalette(RColorBrewer::brewer.pal(9,"Blues"))(30),
                        name = "log2(q)",
)


```

## GO enrichment

```{r GO, echo=TRUE, fig.height = 12, fig.width = 7, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}
dotplot(new_coimg.upregGO, split="ONTOLOGY", showCategory=10) + facet_grid(ONTOLOGY~., scale="free")
dotplot(new_coimg.downregGO, split="ONTOLOGY", showCategory=10) + facet_grid(ONTOLOGY~., scale="free")

```

## IFNy CO vs COiMG

```{r enrichment, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}
ctr_CO <- batch.rem3_mod1[,colnames(batch.rem3_mod1) %in% rownames(meta[meta$COiMg %in% "no" & meta$condition %in% "IFNg",])]
ctr_COiMG <- batch.rem3_mod1[,colnames(batch.rem3_mod1) %in% rownames(meta[meta$COiMg %in% "yes" & meta$condition %in% "IFNg",])]
cor(ctr_CO,ctr_COiMG)

df1 <- data.frame('CO'=rowMeans(ctr_CO))
df2 <- data.frame('COiMg'=rowMeans(ctr_COiMG))

df.plot <- cbind(df1,df2)

quantile(as.matrix(df1-df2))
rownames(df.plot)[df1-df2 >= 1 | df1-df2 <= -4]

Gene <- ifelse(rownames(df.plot) %in% rownames(df.plot)[df1-df2 >= 1 | df1-df2 <= -4], rownames(df.plot), "")

ggplot(df.plot, aes(x=CO,y=COiMg))+
  geom_point() +
  labs(title = "Expression overlap CO vs COiMg under IFNy stimulation",
       x = "Standardized CO counts",
       y = "Standardized COiMg counts") +
  geom_smooth(method=lm,  linetype="dashed",
             color="darkred", fill="blue", se=TRUE) +
  geom_point(shape=18, color="grey")+
  stat_cor(method = "pearson", label.x = 0, label.y = 20)+
  geom_text_repel(data = df.plot, aes(x = CO, y = COiMg, label = Gene), vjust = 3, size = 3) +
  theme_minimal()

```

# Stimulation (IFNy) samples

## PCA plots
```{r pca.2, echo=TRUE, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}

pca.ifny[pca.ifny$Line %in% 'MSN38',]$Line <- 'MSN-38'
pca.ifny[pca.ifny$Line %in% 'MSN9',]$Line <- 'MSN-9'
pca.ifny[pca.ifny$Line %in% 'WTC11',]$Line <- 'WTC-11'


PCAplot(pca.ifny, "condition", Shape = "Line", PoV.df=PoV.ifny, pc.1 = 1, pc.2 = 2, colors = c('lightgrey','#98FB98')) #M30 & M33 are outleirs, let's remove them for now

```

## Volcano plot

```{r volcano.stim, echo=TRUE,  fig.height = 8, fig.width = 6, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}
new_IFNy$symbol <- rownames(new_IFNy)
top30.new_IFNy <- c(rownames(new_IFNy[order(new_IFNy$log2FoldChange),][new_IFNy[order(new_IFNy$log2FoldChange),]$padj < 0.05,][1:15,]),
                    rownames(new_IFNy[order(new_IFNy$log2FoldChange, decreasing = T),][new_IFNy[order(new_IFNy$log2FoldChange, decreasing = T),]$padj < 0.05,][1:15,]))

volcano_plot(data.frame(new_IFNy), title = 'ctr vs IFNy',
             annotate_by=top30.new_IFNy, ymax1 = 60, ymax2 = 61, xmax1 = -5, xmax2 = 10) + 
  scale_x_continuous(breaks = c(-5,-3,-1, 0,1,3,5),
                     labels = c("-5",
                                '-3',
                                "-1",
                                "0",
                                "1",
                                '3',
                                "5"))


```

## Boxplot genes of interest

```{r dea.stim, echo=TRUE, fig.height = 4, fig.width = 5, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}
meta.CO[meta.CO$Line %in% 'MSN38',]$Line <- 'MSN-38'
meta.CO[meta.CO$Line %in% 'MSN9',]$Line <- 'MSN-9'
meta.CO[meta.CO$Line %in% 'WTC11',]$Line <- 'WTC-11'

boxplot.g_ifny_down <- boxplot.g_ifny[boxplot.g_ifny$Sample %in% rownames(meta.CO[!meta.CO$condition %in% 'LPS',]),]
boxplot.g_ifny_down <- boxplot.g_ifny_down[,colnames(boxplot.g_ifny_down) %in% c("CLDN5","IGF1","Stimulation","Sample")]
boxplot.g_ifny_down <- gather(boxplot.g_ifny_down, key = "Gene", value = "Expression", -Sample, -Stimulation)

boxplot.g_ifny_up <- boxplot.g_ifny[boxplot.g_ifny$Sample %in% rownames(meta.CO[!meta.CO$condition %in% 'LPS',]),]
boxplot.g_ifny_up <- boxplot.g_ifny_up[,!colnames(boxplot.g_ifny_up) %in% c("CLDN5","IGF1")]
boxplot.g_ifny_up <- gather(boxplot.g_ifny_up, key = "Gene", value = "Expression", -Sample, -Stimulation)


ggplot(boxplot.g_ifny_down,aes(x = Gene, y = Expression, fill = Stimulation)) +
  geom_boxplot(position = position_dodge(width = 0.8), width = 0.3, alpha = 0.8, outlier.shape = NA)  +
  geom_point(aes(fill = Stimulation), size = 3, shape = 21, position = position_jitterdodge(dodge.width = 0.8, jitter.height = 0.2, jitter.width = 0.01)) +
  theme_minimal() + 
  scale_fill_manual(values=c("white","black")) +
  labs(y = "Standardized expression")
ggplot(boxplot.g_ifny_up,aes(x = Gene, y = Expression, fill = Stimulation)) +
  geom_boxplot(position = position_dodge(width = 0.8), width = 0.3, alpha = 0.8, outlier.shape = NA)  +
  geom_point(aes(fill = Stimulation), size = 3, shape = 21, position = position_jitterdodge(dodge.width = 0.8, jitter.height = 0.2, jitter.width = 0.01)) +
  theme_minimal() + 
  scale_fill_manual(values=c("white","black")) +
  labs(y = "Standardized expression")


```


## Top 40 DEGs heatmap

```{r heatmap.stim, echo=TRUE, fig.height = 10, fig.width = 6, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}
ra.ifny <- HeatmapAnnotation(
  Condition = meta.CO[!meta.CO$condition %in% 'LPS',]$condition,
  col = list(
    Condition = c("IFNg"='#98FB98','ctrl'='white')),
  show_annotation_name = T,
  show_legend = T)

top50.new_IFNy <- c(rownames(new_IFNy[order(new_IFNy$log2FoldChange),][new_IFNy[order(new_IFNy$log2FoldChange),]$padj < 0.05,][1:25,]),
                    rownames(new_IFNy[order(new_IFNy$log2FoldChange, decreasing = T),][new_IFNy[order(new_IFNy$log2FoldChange, decreasing = T),]$padj < 0.05,][1:25,]))

hm_counts.ifny <- t(scale(t(batch.ifny[rownames(batch.ifny) %in% top50.new_IFNy,])))

ComplexHeatmap::Heatmap(hm_counts.ifny,
                        cluster_rows = T,
                        cluster_columns = T,
                        show_row_dend = T,
                        show_column_dend = T,
                        show_row_names = F,
                        show_column_names = F,
                        top_annotation = ra.ifny,
                        right_annotation = rowAnnotation(
                          text = anno_text(rownames(hm_counts.ifny), rot = 0, just = "left",gp=gpar(fontsize=8))),
                        col = colorRampPalette(rev(RColorBrewer::brewer.pal(9,"RdBu")))(10),
                        name = "Z-score expression")

```

## Enrichment heatmap (gene sets of interest)
```{r heatmap.stim2, echo=TRUE, fig.height = 6, fig.width = 3, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}

ComplexHeatmap::Heatmap(DEG.enrich_ifny[!rownames(DEG.enrich_ifny) %in% 'Lipid metabolism',],
                        cluster_rows = F,
                        cluster_columns = F,
                        show_row_dend = F,
                        show_column_dend = F,
                        show_row_names = T,
                        col = colorRampPalette(RColorBrewer::brewer.pal(9,"Blues"))(30),
                        name = "log2(q)",
)


```

## NDD gene set enrichment in IFNy
```{r ifny.ndd, echo=TRUE, fig.height = 6, fig.width = 8, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}

cgenes <- readxl::read_xlsx('/Users/kubler01/Documents/R projects/gene lists/ndd/Gene lists for organoid paper.xlsx', col_names = T)
m <- data.frame(t(toupper(na.omit(cgenes$`ADHD GWAS associated genes through FUMA`))[toupper(na.omit(cgenes$`ADHD GWAS associated genes through FUMA`)) %in% rownames(batch.ifny)]), check.names = F)
for(i in names(cgenes)[-1]){
  k <- na.omit(cgenes[i])
  s <- k[as.matrix(k)[,1] %in% rownames(batch.ifny),]
  m <- rbind.fill(data.frame(m),data.frame(t(s)))
}
rownames(m) <- names(cgenes)
m <- t(m)

#extract DEGs
DEGs.ifny <- list("Up-regulated"=na.omit(all_res$IFNg$DEGs[,1]),"Down-regulated"=na.omit(all_res$IFNg$DEGs[,2]))


#gene set enrichment analysis:
ndd.enrich <- matrix(nrow=2*length(colnames(m)), ncol=6)
ndd.rep <- NULL
for (i in colnames(m))
  ndd.rep <- c(ndd.rep, paste(i, "_up", sep=""), paste(i, "_down", sep=""))
rownames(ndd.enrich) <- ndd.rep
colnames(ndd.enrich) <- colnames(GSEA.byMod(mod.gl=DEGs.ifny, m[,1], universe=35324))


for (x in colnames(m)){
  ndd.enrich[gsub("*_up","",gsub("*_down","",rownames(ndd.enrich))) %in% x,] <- as.matrix(GSEA.byMod(mod.gl=DEGs.ifny, na.omit(as.matrix(m)[,x]), universe=na.omit(rownames(res))))
}

ndd.enrich <- data.frame(ndd.enrich)
ndd.enrich$log.q <- -log2(ndd.enrich[,6])

#DEG.enrich.res[is.infinite(DEG.enrich.res$log.q),]$log.q <- 1
ndd.enrich.ress <- cbind("Upregulated"=ndd.enrich[c(seq(1, length(rownames(ndd.enrich))-1, by = 2)),][,7],
                    "Downregulated"=ndd.enrich[c(seq(2, length(rownames(ndd.enrich)), by = 2)),][,7])
rownames(ndd.enrich.ress) <- colnames(m)

ComplexHeatmap::Heatmap(ndd.enrich.ress,
                        cluster_rows = F,
                        cluster_columns = F,
                        show_row_dend = F,
                        show_column_dend = F,
                        show_row_names = T,
                        col = colorRampPalette(RColorBrewer::brewer.pal(9,"Blues"))(30),
                        name = "log2(q)",
)


```

## GO enrichment

```{r GO.stim, echo=TRUE, fig.height = 12, fig.width = 7, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}
dotplot(new_IFNy.upregGO, split="ONTOLOGY", showCategory=10) + facet_grid(ONTOLOGY~., scale="free")
dotplot(new_IFNy.downregGO, split="ONTOLOGY", showCategory=10) + facet_grid(ONTOLOGY~., scale="free")

```

## iMG vs ocMG

```{r enrichment.stim, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}
#==========iMG - COiMG correlation===============+#
#integrate iMG with COiMG
load('/Users/kubler01/Documents/R projects/bulk-org/data/Psamples_gene_matrix.RData')
genes_counts
meta_p <- data.frame(readxl::read_xlsx('/Users/kubler01/Documents/R projects/bulk-org/meta_P.xlsx'))
counts_p <- genes_counts[,colnames(genes_counts) %in% meta_p$ID]
counts_P.filt <- counts_p[rownames(counts_p) %in% counts_protcod2,]
rownames(counts_P.filt) <- make.unique(protcod_genes_reduced[,2])
#filter genes on expression
gene_names <- rownames(counts_P.filt)
cpm = edgeR::cpm(counts_P.filt)
median_genes_cpm <- enframe(rowMedians(as.matrix(counts_P.filt)), name = "Gene", value = "median_cpm")
median_genes_cpm <- cbind(median_genes_cpm, gene_names)
keep.exp <- dplyr::filter(median_genes_cpm, median_cpm > 1)
keep.exp <- keep.exp$gene_names

counts_P.filt2 <- counts_P.filt[keep.exp,]


dds.P <- DESeqDataSetFromMatrix(countData = round(counts_P.filt2),
                                   colData = meta_p,
                                   design = ~ RIN + `X260.280` + `X260.230` + type) # RIN + Line + Age + `260/230` + `260/280` + 

vsd.p <- vst(dds.P)
transformed.p <- data.frame(assay(vsd.p), check.names=F)


batch.p <- removeBatchEffect(transformed.p, 
                                covariates=as.matrix(cbind(
                                  meta_p$`X260.280`,
                                  meta_p$`X260.230`,
                                  meta_p$RIN)),
                                design=model.matrix(~ meta_p$type))


iMG <- batch.p[,colnames(batch.p) %in% meta_p[meta_p$type %in% 'iMG',]$ID]
ocMG <- batch.p[,colnames(batch.p) %in% meta_p[meta_p$type %in% 'ocMG',]$ID]
cor(iMG,ocMG)

df1 <- data.frame('iMG'=rowMeans(iMG))
df2 <- data.frame('ocMG'=rowMeans(ocMG))

df.plot <- cbind(df1,df2)

quantile(as.matrix(df1-df2))
rownames(df.plot)[df1-df2 >= 0.5 | df1-df2 <= -1]

Gene <- ifelse(rownames(df.plot) %in% rownames(df.plot)[df1-df2 >= 0.5 | df1-df2 <= -1], rownames(df.plot), "")

ggplot(df.plot, aes(x=iMG,y=ocMG))+
  geom_point() +
  labs(title = "Expression overlap iMG and ocMG",
       x = "Standardized iMG counts",
       y = "Standardized ocMG counts") +
  geom_smooth(method=lm,  linetype="dashed",
             color="darkred", fill="blue", se=TRUE) +
  geom_point(shape=18, color="grey")+
  stat_cor(method = "pearson", label.x = 0, label.y = 20)+
  geom_text_repel(data = df.plot, aes(x = iMG, y = ocMG, label = Gene), vjust = 3, size = 3) +
  theme_minimal()


```



# IFNy in COiMG

## Volcano plot

```{r ifny.coimg, echo=TRUE, fig.height = 9, fig.width = 6, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}
load("/Users/kubler01/Documents/R projects/bulk-org/01232024_figures AB_ifny in coimg.RData")


volcano_plot(data.frame(new_coimg_ifny), title = 'ctr vs IFNy in COiMg',
             annotate_by=top30.new_coimg_ifny, ymax1 = 80, ymax2 = 81, xmax1 = -5, xmax2 = 15) + 
  scale_x_continuous(breaks = c(-5,-3,-1, 0,1,3,5),
                     labels = c("-5",
                                '-3',
                                "-1",
                                "0",
                                "1",
                                '3',
                                "5"))


```


## lFC correlation plot
```{r ifny.coimg_cor, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center", message = FALSE, warning=FALSE, eval=TRUE}

lfc.plot <- data.frame('Gene'=new_IFNy$symbol,'CO'=new_IFNy$log2FoldChange,'COiMg'=new_coimg_ifny[rownames(new_coimg_ifny) %in% rownames(new_IFNy),]$log2FoldChange)
rownames(lfc.plot) <- new_IFNy$symbol

Gene <- ifelse(rownames(lfc.plot) %in% rownames(lfc.plot)[lfc.plot$CO-lfc.plot$COiMG >= 4 | lfc.plot$CO-lfc.plot$COiMG <= -4], rownames(lfc.plot), "")

ggplot(lfc.plot, aes(x=CO,y=COiMg))+
  geom_point() +
  labs(title = "lFC correlation in CO & COiMg under IFNy stimulation",
       x = "lFC IFNy in CO",
       y = "lFC IFNy in COiMg") +
  geom_smooth(method=lm,  linetype="dashed",
             color="darkred", fill="blue", se=TRUE) +
  geom_point(shape=18, color="grey")+
  stat_cor(method = "pearson", label.x = 0, label.y = 20)+
  geom_text_repel(data = lfc.plot, aes(x = CO, y = COiMg, label = Gene), vjust = 3, size = 3) +
  theme_minimal()
```